Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-19 aistats.tex
Simplify multivariate bound

\begin{theorem}[Multivariate MLE Upper Bound]
	The MLE of  the covariance matrix of $X_i\sim\cN(0,\mu_*)$ is $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top \in\real^{d\times d}$ with $p = \frac{d(d+1)}{2} $ degrees of freedom. 
	The expected KL between $\mu_*$ and $\hat\mu_n$ is infinite when $n\leq d+1$ and otherwise upper bounded by
	\alignn{
	 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
 	\ , \forall n >d+1 \;.
}
This bound is asymptotically tight.
\end{theorem}
\begin{proof}
The entropy of $X$ is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, whose gradient is the negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$.
The associated Bregman divergence is
\alignn{
	\bregmanconj(\mu_*; \hat \mu_n) 
	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
}
Thanks to the linearity of the trace, the expectation becomes
\alignn{
	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
}

When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
then the mean parameter "ratio" is
\aligns{
	\mV = n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top
}
but $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$.
Recall that $\E[\mV] = n \mI$.
Thanks to $\log\det$ being a sufficient statistic of the Wishart, and the natural to mean parameter formula~\eqref{eq:mirror-map}, we have a closed form for the expected log-determinant of a Wishart 
\aligns{
	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
}
where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
The expectation of an inverse Wishart  is straightforward to compute from the density and the log-partition function
\begin{equation}
	\E[\mV^{-1}] = \begin{cases}
		+\infty \text{ if } n\leq d+1 \\
		 \frac{\mI}{n - d - 1} \text{ otherwise,} 
	\end{cases}
\end{equation}
which proves the infinite part of the statement.
We now consider only $n>d+1$.
We deduce that $\Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}$.
Putting it all together, we get the following closed form for the expectation of the KL
 \begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
 \end{align}

The complicated term is the $\psi_d$ function.
We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
	\label{eq:psid_bound}
\end{align}
We can bound the harmonic sum thanks to the typical bound
\aligns{
	\inv{2n +1}
	\leq H_n - \log(n) - \gamma \leq 
	\inv{2n - 1}
}
where $\gamma$ is the Euler constant. Then
\begin{align}
	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
	&= H_{n-d} - H_n \\
	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
\end{align}
Plugging this back into~\eqref{eq:psid_bound} yields
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1} 
\end{align}
where we emphasized that the sum runs up to $d$.
Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
\begin{align}
	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
	\leq \sum_{i=0}^{d} - \frac{i}{n}
	 = -\frac{d(d+1)}{2n}
\end{align}
which yields the final rate
\begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
 	\ , \forall n >d+1 \;.
\end{align}
where $p = d(d+1)/2$.
\end{proof}

>>>>

\begin{theorem}[Multivariate MLE Upper Bound]
	The MLE of  the covariance matrix of $X_i\sim\cN(0,\mu_*)$ is $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top \in\real^{d\times d}$ with $p = \frac{d(d+1)}{2} $ degrees of freedom. 
	The expected KL between $\mu_*$ and $\hat\mu_n$ is infinite when $n\leq d+1$ and otherwise upper bounded by
	\alignn{
	 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{p}{2n} + \frac{p(d+2)}{n (n - d -1)} 
 	\ , \forall n >d+1 \;.
}
This bound is asymptotically tight.
\end{theorem}
\begin{proof}
The entropy of $X$ is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, whose gradient is the negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$.
The associated Bregman divergence is
\alignn{
	\bregmanconj(\mu_*; \hat \mu_n) 
	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
}
Thanks to the linearity of the trace, the expectation becomes
\alignn{
	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
}

When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
then we define the mean parameter "ratio" as
\aligns{
	\mV &:= n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top 
}
such that $\Tr(\mu_* \hat \mu_n^{-1})  = n \Tr(\mV^{-1})$.
But $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$.
Recall that $\E[\mV] = n \mI$.
Thanks to $\log\det$ being a sufficient statistic of the Wishart, and the natural to mean parameter formula~\eqref{eq:mirror-map}, we have a closed form for the expected log-determinant of a Wishart 
\aligns{
	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
}
where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
The expectation of an inverse Wishart  is straightforward to compute from the density and the log-partition function
\begin{equation}
	\E[\mV^{-1}] = \begin{cases}
		+\infty \text{ if } n\leq d+1 \\
		 \frac{\mI}{n - d - 1} \text{ otherwise,} 
	\end{cases}
\end{equation}
which proves the infinite part of the statement.
We now consider only $n>d+1$.
We deduce that 
\alignn{
	\Tr(\expect{\mu_* \hat \mu_n^{-1}} - d
	= n\Tr(\E[\mV^{-1}]) - d 
	= \frac{nd}{n - d - 1} -d
	= \frac{d (d+1}{n -d -1}
	= \frac{2p}{n -d -1}
}
Putting it all together, we get the following closed form for the expectation of the KL
 \begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	= \half \left ( \frac{2p}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
 \end{align}

The complicated term is the $\psi_d$ function.
We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
	\label{eq:psid_bound}
\end{align}
We can bound the harmonic sum thanks to the typical bound
\aligns{
	\inv{2n +1}
	\leq H_n - \log(n) - \gamma \leq 
	\inv{2n - 1}
}
where $\gamma$ is the Euler constant. Then
\begin{align}
	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
	&= H_{n-d} - H_n \\
	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)}
\end{align}
Plugging this back into~\eqref{eq:psid_bound} yields
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)} 
\end{align}
where we emphasized that the sum runs up to $d$.
Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
\begin{align}
	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
	\leq \sum_{i=0}^{d} - \frac{i}{n}
	 = -\frac{d(d+1)}{2n} = -\frac{p}{n} \; .
\end{align}
Putting it all together yields
\begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{(d + 1)}{(2n+1)(2(n - d) - 1)} \\
 	&\leq p \frac{n + d +1}{2 n (n-d-1)}  + \frac{(d + 1)}{2n.2 (n - d - 1)}\\
 	&= \frac{p}{2n} + \frac{2 p(d+1)}{2 n (n-d-1)} + \frac{p/d}{2n (n - d - 1)} \\
 	&\leq \frac{p}{2n} + \frac{p(d+2)}{n (n - d -1)} 
 	\ , \forall n >d+1 \;.
\end{align}
where $p = d(d+1)/2$ and the last inequality used $\inv{2d}\leq 1$. 
\end{proof}
